The role of aneuploidy in the evolution of cancer drug resistance

### 2023-07-20 EvolutionaryRescue1.tex
revised evol time section

In our model, \emph{evolutionary rescue} occurs when resistant cells appear and fixate ($m_t \gg 1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.

To estimate the rescue probability $\presc$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
Thus, the rescue probability is given by 
\begin{subequations}
\begin{align}
\presc &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
& \approx 1-\e^{-Np_w} , \label{rescue_prob2}
\end{align}
\end{subequations}
where $p_w$ is the probability that the lineage of a single wild-type cell avoids extinction, and the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not necessarily $N p_w$) is small.

The probability for evolutionary rescue $\presc$  (\cref{rescue_prob2}) is $\presc \approx N p_w$ when $N<1/p_w$ and $\presc \approx 1$ when $N > 1/p_w$. Therefore, $N^*=1/p_w$ is the threshold tumor size for rescue above which rescue is very likely (\Cref{rescue_prob_N}). % TODO add 1/p_w to figure as vertical line
Therefore, we can write
\begin{equation}
\presc = 1-e^{-N/N^*} .
\end{equation}
In the Appendix below, we use the theory of multi-type branching processes to find approximate expressions \cref{survprobwapprox1,survprobwinitial,scenario3} for $p_w$ in different regimes. 
Substituting these  into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
\\

In the following, an important quantity is $T^* = (4 v \lambda_a \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u T^*< 1$), or if aneuploidy is rare ($u < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that 
\begin{equation}
N_m^* \approx \frac{\abs{\Delta_w}}{v} \cdot \frac{\lambda_m}{\Delta_m} .
\end{equation}
Otherwise, aneuploidy is frequent enough ($u > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of the tumor. 
The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time,
\begin{equation}\label{rescue_prob_approx}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v} \cdot \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 ,\\ 
  %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{aligned}
\end{equation}
%The second case has a finer approximation by $N^*_a \approx \frac{\abs{\Delta_w}}{u} \cdot \frac{2\lambda_a}{\Delta_a + 1/T^*}$.
Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).
\\

The ratio of threshold tumor size for rescue by aneuploidy ($u$ is high) or by direct mutation ($u$ is low) is
\begin{equation}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{u}\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
%  \sqrt{\frac{\Delta_a}{u} \cdot v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \cdot \left(u \frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
In all cases, the threshold size ratio rapidly decreases with the aneuploidy rate $u$ and the aneuploid growth rate $\Delta_a$ increase, although when $\Delta_a$ is close to zero, it has XXX

In the first case, $\abs{\Delta_a}/u$ is  the ratio of the expected time for an aneuploid lineage appears, $1/u$ and the expected time until that lineage disappears, $1/\Delta_a$.
In the third case, $v \frac{\Delta_m}{\lambda_m} \cdot \left(u \frac{\Delta_a}{\lambda_a}\right)^{-1}$ is the ratio of the rate of formation of resistant mutants that avoid extinction and partially resistant aneuploid that avoid extinction.
In the second case, $\frac{1}{u}\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u} \cdot v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third cases.

Interestingly, if we increase $\lambda_a$ and $\mu_a$ such that $\Delta_a$ remains constant, $T^*$ decreases, and we are in the middle case. In this case, increasing $\lambda_a$ should also increase the mutation rate in aneuploid cells, $v$, so overall $d \lambda_a T^*/d\lambda_a = 0$ and $N_a^*$ is unaffected by increasing $\lambda_a$. 
% TODO PLUG IN REAL VALUES

% TODO move this to appendix
%Substituting these  into \cref{rescue_prob2}, we find that the rescue probability can be approximated by
%\begin{equation}\label{rescue_prob_approx}
%\begin{aligned}
%&\presc \approx \\
%  &\begin{cases}
%  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
%  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
%   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
%   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
%   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
%   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
%  \end{cases}
%\end{aligned}
%\end{equation}

@@

In our analysis we used branching processes, which assume that growth (division and death) are density-independent. However, growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density. 
We therefore performed stochastic simulations of a logistic growth model with carrying capacity $K$ (see Methods). 
We find that our approximations agree with results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{rescue_prob_N}).

@@

In the above we assumed that upon beginning of drug therapy, the initial tumor consisted entirely of wildtype cells.
However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells REF), which are likely to have a deleterious effect, $s$ (REF). % TODO refs
But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f=\tilde{u}/s$ of aneuploid cells in the population (REF).

Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo aneuploidy, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that 
\begin{equation}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}} .
\end{equation}

Therefore, standing genetic variation will drive adaptation to the drug if $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
In contrast, de-novo aneuploids will have a stronger effect on adaptation if the cost $s$ is large, the drug is weak ($\Delta_w$ is small) or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).

@@

%Plugging in our approximations from \cref{rescue_prob}, we have
%\begin{equation}\label{ratiorescue}
%H = \begin{cases}
%\frac{1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right]}{1-\exp\left[\frac{vN\Delta_m}{\left(\Delta_w-v\right)\lambda_m}\right]} ,&
%4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
%\frac{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right]}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)}, ,&
%\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
%\frac{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]}{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\right]} ,&
%\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2. % TODO this term can be simplified?
%  \end{cases}
%\end{equation}
%
%We find that the rescue ratio increase with the aneuploidy growth rate $\Delta_a$, because the better aneuploid cells are in growth, the better they are at rescuing the population (when they provide partial resistance) or delaying the extinction of the population (when they provide tolerance). 
%However, the rescue decreases with the wildtype growth rate $\Delta_w$, because the better the wildtype is at growth, the less is depends on aneuploidy for rescue or delay, and the more likely it is to directly produce mutant cells, rather than relying on aneuploid cells for producing mutant cells (\Cref{rescue_ratio}).
%The effect of the initial tumor size $N$ is the similar to that of the wildtype growth rate. 
%Importantly, in large tumors, the ratio converges to unity, that is, aneuploidy does not affect the probability for evolutionary rescue.

@@

Even when evolutionary rescue occurs, it may take a long time.
We therefore wish to estimate the mean waiting time for rescue and the effect aneuploidy may have on it. 
We calculate the mean time for the appearance of the first mutant that rescues the cell population.
This can occur either through the evolutionary trajectory $wildtype \rightarrow mutant$ or through the trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$.
We start with the former.

The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$,
where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
Note that 
\begin{equation}\label{integralR}
\int_0^{t}{R(z)\d z} = 
u p_a N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
u p_a N t,
\end{equation}
by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
The probability density function of $T_1$ is thus
$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
Therefore, the probability density function of the conditional random variable $(T_1 \mid T_1 < \infty)$ is
$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
\\

We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
\begin{equation}\label{meantime1}
\begin{aligned}
\tau_1 =
\int_{0}^{\infty}{t f_1(t) \d t} = 
\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
\end{aligned}
\end{equation}
after applying integration by parts.
Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, 
\begin{align}\label{limitapprox}
\tau_1 = 
%\frac{\int_0^\infty\exp\left(-\int_0^\tau R(z)\d z\right)\d t}{1-\left(1-p_w\right)^N} = 
\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t}{1-\left(1-p_w\right)^N} \approx
\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx \\
\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
\frac{1+\e^{-Np_w}}{u p_a N} ,
\end{align}
where we use the approximations 
$\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2)$ and $(1-\e^{-Np_w})^{-1} \approx 1+\e^{-Np_w}$ and integrate the exponent.
\Cref{MeanTimeGrowthAneuploidyPlot}B show the agreement between this approximating and simulation results for intermediate and large tumor sizes.
\\

We wish to obtain a simpler formula for $\tau_2$, similar to \cref{limitapprox}.
We thus have the following expansions,
%\begin{align*}
%\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3) ,\\
%\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3) ,
%\end{align*}
%which allow us to write
%\begin{equation*}
%\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a} \approx \frac{\left(\Delta_w-\Delta_a\right)\tau^2}{2}.
%\end{equation*}
%As a result, the integrand in \cref{meantimet2} can be written as:
%\begin{align*}
%&\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\approx \\
%& \exp\left(-uvNp_m\tau^2-vNp_m\tau\right) =
%\exp\left(\frac{vNp_m}{2}\right)\exp\left[-\frac{uvNp_m}{2}\left(\tau+\frac{1}{u}\right)\right].
%\end{align*}
%Consequently, the mean conditional time $\tau_2$ is approximated by
%\begin{align}\label{limitapprox2}
%\tau_2 \approx
%\frac{\left(1+\exp\left(-Np_w\right)\right) \exp\left(\frac{vNp_m}{2u}\right) \erfc\left(\sqrt{\frac{vNp_m}{2u}}\right)}{\sqrt{\frac{2uvNp_m}{\pi}}},
%\end{align}
%where $\erfc(z)$ is the complementary error function. 
%
%If we apply linear approximations,
\begin{align*}
\frac{\e^{\Delta_w t}-1}{\Delta_w}&=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w} = t+O(t^2),\\
\frac{\e^{\Delta_a t}-1}{\Delta_a}&=\frac{1+\Delta_a t+O(t^2)-1}{\Delta_a} = t+O(t^2),
\end{align*}
which we use to derive a first-order approximation for $\tau_2$,
\begin{align}\label{limitapprox3}
\tau_2 \approx
\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u N p_m t} \d t = 
\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
\end{align}

\Cref{MeanTimeGrowthAneuploidyPlot}A shows that the approximation \cref{limitapprox3} has a good fit with simulation results for large initial wildtype population size ($N > 2\cdot10^7$). An approximation that uses the second-order terms, $\frac{\Delta_w^2 t^2}{2}$ and $\frac{\Delta_a^2 t^2}{2}$, does not perform better (\Cref{MeanTimeGrowthAneuploidyPlot}A).
\\

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Effect of aneuploidy on mean evolutionary rescue time}

Next, we wish to find the ratio of the mean rescue time with and without aneuploidy ($u>0$ and $u=0$, respectively),
\begin{align}\label{eqMeanTimeRatioInitialPopulationSize}
\frac{\tau_2}{\tau_1}  
%\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w t}-1}{\Delta_w}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right] \d t}{\int_0^\infty\e^{-uNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t} \cdot \frac{1-\left(1-p_w(u=0)\right)^N}{1-\left(1-p_w(u>0)\right)^N}\\
= H^{-1} \cdot \frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right] \d t}{\int_0^\infty\e^{-vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t},
\end{align}
where $H$ is the ratio of the rescue probability with and without aneuploidy, see \cref{ratiorescueexact}.
\Cref{MeanTimeRatioInitialPopulationSize} shows how this ratio, $\tau_2/\tau_1$, changes as a function of the initial tumor size, $N$, for different aneuploid growth rates, $\Delta_a$.

@@

We have modeled a tumor--a population of cancer cells--exposed to drug therapy that causes the population to decline in size towards potential extinction.
The cancer cell population can be "evolutionary rescued", or escape extinction, by two paths. In the direct path, a sensitive cell acquires a mutation that confers resistance that allows it to rapidly grow. In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the effect of the drug, and then an aneuploid cell acquires a mutation that confers resistance (\Cref{figureAneuploidy}).

Using multitype branching processes, we derived the probability of evolutionary rescue of the population of cancer cells under different scenarios for the effect of aneuploidy, ranging from tolerance to partial resistance.
We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{rescue_prob}).
Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_w$, the mutation rate $v$, and the aneuploidy rate $u$.

@@

\appendix
\begin{appendices}
\renewcommand{\theequation}{\thesection\arabic{equation}}
\counterwithin*{equation}{section}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Survival probability of a single lineage}

>>>>

In our model, \emph{evolutionary rescue} occurs when resistant cells appear and fixate ($m_t \gg 1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.
We assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
We therefore define $p_w$ as the probability that a lineage starting from a single wildtype cell avoids extinction by acquiring drug resistance.
Thus, $N^*=1/p_w$ is the threshold tumor size above which evolutionary rescue is very likely, and the rescue probability is given by 
\begin{equation} \label{eq:rescue_prob} 
\presc = 
1-\left(1-p_w\right)^N \approx
1-\e^{-Np_w} = 
1-e^{-N/N^*} .
\end{equation}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not necessarily $N p_w$) is small.
Indeed, when $N<1/p_w$, then the probability for evolutionary rescue is $\presc \approx N p_w$  and when $N > 1/p_w$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size. 
\\

In the Appendix, we use the theory of multi-type branching processes to find approximate expressions \cref{eq:survprobwapprox1,eq:survprobwinitial,eq:scenario3} for $p_w$ in different regimes. 
Substituting these  into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
For these approximations, an important quantity is $T^* = (4 v \lambda_a \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u T^*< 1$), or if aneuploidy is rare ($u < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that 
\begin{equation} \label{eq:N_m}
N_m^* \approx \frac{\abs{\Delta_w}}{v} \cdot \frac{\lambda_m}{\Delta_m} .
\end{equation}
Here, $\abs{\Delta_w}/v$ is the ratio of XXX.

Otherwise, aneuploidy is frequent enough ($u > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of drug resistance. 
The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v} \cdot \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 ,\\ 
  %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{aligned}
\end{equation}
%The second case has a finer approximation by $N^*_a \approx \frac{\abs{\Delta_w}}{u} \cdot \frac{2\lambda_a}{\Delta_a + 1/T^*}$.
Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).

Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
\begin{equation} \label{eq:N_ratio}
\frac{N^*_a}{N^*_m} \approx \begin{cases}
    \frac{\abs{\Delta_a}}{u} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{u}\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
%  \sqrt{\frac{\Delta_a}{u} \cdot v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \cdot \left(u \frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
In all cases, the effect of aneuploidy increases (i.e., the threshold size ratio decreases) when the aneuploidy rate $u$ increases. Increasing the aneuploid growth rate $\Delta_a$ also leads to an increased role of aneuploidy, but only in the first and third case.

In the first case, $\abs{\Delta_a}/u$ is  the ratio of the expected time for an aneuploid lineage to appear, $1/u$, and the expected time until that lineage disappears, $1/\Delta_a$.
In the third case, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of formation of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
In the second case, $\frac{1}{u}\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u} \cdot v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third cases.

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, and therefore to the second case. In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v$ in aneuploid cells, as mutations mostly occur during division , so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploids rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation.
% TODO PLUG IN REAL VALUES

@@

In our analysis we used branching processes, which assume that growth (division and death) is density-independent. However, growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density. 
We therefore performed stochastic simulations of a logistic growth model with carrying capacity $K$ (see Methods). 
We find that our approximations agree with results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{rescue_prob_N}).

@@

In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells REF), which are likely to have a deleterious effect in the absence of the drug, $s$ (REF). % TODO refs
But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f \approx \tilde{u}/s$ of aneuploid cells in the population.

Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that 
\begin{equation}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}} .
\end{equation}

Therefore, standing genetic variation will drive adaptation to the drug if $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
In contrast, de-novo aneuploids will have a stronger effect on adaptation if the aneuploidy cost $s$ is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).

@@

%Plugging in our approximations from \cref{eq:rescue_prob}, we have
%\begin{equation}\label{ratiorescue}
%H = \begin{cases}
%\frac{1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right]}{1-\exp\left[\frac{vN\Delta_m}{\left(\Delta_w-v\right)\lambda_m}\right]} ,&
%4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
%\frac{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right]}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)}, ,&
%\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
%\frac{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]}{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\right]} ,&
%\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2. % TODO this term can be simplified?
%  \end{cases}
%\end{equation}
%
%We find that the rescue ratio increase with the aneuploidy growth rate $\Delta_a$, because the better aneuploid cells are in growth, the better they are at rescuing the population (when they provide partial resistance) or delaying the extinction of the population (when they provide tolerance). 
%However, the rescue decreases with the wildtype growth rate $\Delta_w$, because the better the wildtype is at growth, the less is depends on aneuploidy for rescue or delay, and the more likely it is to directly produce mutant cells, rather than relying on aneuploid cells for producing mutant cells (\Cref{rescue_ratio}).
%The effect of the initial tumor size $N$ is the similar to that of the wildtype growth rate. 
%Importantly, in large tumors, the ratio converges to unity, that is, aneuploidy does not affect the probability for evolutionary rescue.

@@

Even when evolutionary rescue occurs and leads to recurrence of the tumor, it may take a long time.
The overall expected recurrence time can be estimated by adding two terms: the mean waiting time for evolutionary rescue--the appearance of a resistant lineage that avoid extinction--and the expected time for proliferation of that lineage back to the original tumor size, $N$.

\paragraph{Evolutionary rescue time.}
In Appendix C we derive an approximation for $\tau_1$, the expected rescue time without aneuploidy ($u=0$), and $\tau_2$, the expected rescue time with aneuploidy ($u>0$),
\begin{align}
\tau_1 &\approx \frac{1+\e^{-Np_w}}{u N p_a} , \label{eq:limitapprox} ,\\
\tau_2 &\approx \frac{1+\e^{-Np_w}}{u N p_m} , \label{eq:limitapprox3} .
\end{align}
\Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between this approximations and simulation results for intermediate and large tumor sizes.

Thus, the ratio of the expected rescue time with and without aneuploidy ($u>0$ and $u=0$, respectively),
\begin{align}\label{eq:eqMeanTimeRatioInitialPopulationSize}
\frac{\tau_1}{\tau_2}  \approx \frac{p_a}{p_m} = 
%&\frac{\lambda_m}{\lambda_a} \; \frac{\Delta_a-v+\sqrt{(\Delta_a - v)^2 + (1/T^*)^2}}{2 \Delta_m} \approx \\
%&\frac{1}{2} \; \frac{\lambda_m}{\Delta_m} \left[ \left(\frac{\lambda_a}{\Delta_a}\right)^{-1} + \sqrt{\frac{\left(\Delta_a\right)^2 + (1/T^*)^2}{\lambda_a^2}}\right] = \\
\frac{1}{2} \; \frac{\lambda_m}{\Delta_m} \left[ \left(\frac{\lambda_a}{\Delta_a}\right)^{-1} + \sqrt{\left(\frac{\lambda_a}{\Delta_a}\right)^{-2} + 4v \left(\lambda_a\frac{\lambda_m}{\Delta_m}\right)^{-1}}\right] .
\end{align}

\Cref{MeanTimeRatioInitialPopulationSize} shows how this ratio, $\tau_1/\tau_2$, changes as a function of the initial tumor size, $N$, for different aneuploid growth rates, $\Delta_a$.

@@

\paragraph{Proliferation time.}

TODO

@@

We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes the population to decline in size towards potential extinction.
In this scenario, the tumor can be "evolutionary rescued", or escape extinction, via two paths. In the direct path, a sensitive cell acquires a mutation that confers resistance that allows it to rapidly grow. In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the effect of the drug, and then an aneuploid cell acquires a mutation that confers resistance (\Cref{figureAneuploidy}).

Using multitype branching processes, we derived the probability of evolutionary rescue of the population of cancer cells under different scenarios for the effect of aneuploidy, ranging from tolerance to partial resistance.
We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}).
Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_w$, the mutation rate $v$, and the aneuploidy rate $u$.

@@

\newpage 
\clearpage

\begin{appendices}
\renewcommand{\theequation}{\thesection\arabic{equation}}
\counterwithin*{equation}{section}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Appendix A: Survival probability of a single lineage}\label{sec:appendix-surv-prob}
